Analyzing Performance Bottleneck in Graph Neural Network Training: An Experimental View

### 2020-07-13 Manuscript/sec2-review-of-graph-neural-network.tex
Update section 2.

\begin{table}
	\hspace{-10em}
    \begin{footnotesize}
        \begin{tabular}{ccp{8em}p{22em}r}
		\toprule
		GNN &
		Type &
		$\Sigma$ &
		$\phi$ &
		Complexity \\ \midrule
		ChebNet \cite{defferrad2016_chebnet} &
		Spectral &
		sum &
		$\boldsymbol{m}_{j, i}^{(k)} = e_{j, i}\boldsymbol{z}_j^{(k-1)}$ &
		$O(d_{in})$ \\
		\textbf{GCN} &
		Spectral &
		sum &
		$\boldsymbol{m}_{j, i}^l = e_{j, i} \boldsymbol{h}_j^l$ &
		$O(d_{in})$ \\
		AGCN &
		Spectral &
		sum &
		$\boldsymbol{m}_{j, i}^l = \tilde{e}_{j, i}^l \boldsymbol{h}_j^l$ &
		$O(d_{in})$ \\
		GraphSAGE &
		Non-spectral &
		mean/LSTM &
		$\boldsymbol{m}_{j, i}^l =  \boldsymbol{h}_j^l$ &
		$O(1)$ \\
		Neural FPs &
		Non-spectral &
		sum &
		$\boldsymbol{m}_{j, i}^l = \boldsymbol{h}_j^l$ &
		$O(d_{in})$ \\
		SSE &
		Recurrent &
		sum &
		$\boldsymbol{m}_{j, i}^l = [\boldsymbol{h}_i^{l} \parallel \boldsymbol{h}_j^l]$ &
		$O(1)$ \\
		\textbf{GGNN} &
		Gated &
		sum &
		$\boldsymbol{m}_{j, i} = \textcolor{blue}{\boldsymbol{W}^l} \boldsymbol{h}_j^l$ &
		$O(d_{in} * d_{out})$ \\
		\textbf{GAT}&
		Attention &
		sum  &
		\begin{scriptsize}
			$\begin{aligned}[t]
			& \alpha_{j, i}^k = \frac {\exp(LeakyReLU(\textcolor{blue}{\boldsymbol{a}^T} [ \textcolor{blue}{\boldsymbol{W}^{l,k}}  \boldsymbol{h}_j^l \parallel \textcolor{blue}{\boldsymbol{W}^{l,k}}  \boldsymbol{h}_j^l] ))} {\sum_{k \in \mathcal{N}(i)}\exp(LeakyReLU(\textcolor{blue}{\boldsymbol{a}^T} [ \textcolor{blue}{\boldsymbol{W}^{l,k}}  \boldsymbol{h}_j^l \parallel \textcolor{blue}{\boldsymbol{W}^{l,k}}  \boldsymbol{h}_k^l] ))} \\
			&  \boldsymbol{m}_{j, i}^l = \parallel_{k=1}^K \delta(\alpha_{j, i}^k \textcolor{blue}{\boldsymbol{W}^{l,k}} \boldsymbol{h}_j^{l})
			\end{aligned}$
		\end{scriptsize} &
		$O(K * d_{in} * d_{out})$ \\
		\textbf{GaAN} &
		Attention &
		sum,max,mean &
		\begin{scriptsize}
			$\begin{aligned}[t]
			&\alpha_{j, i}^k = \frac {\exp(\textcolor{blue}{\boldsymbol{a}^T} [ \textcolor{blue}{\boldsymbol{W}^{l, k}_{xa}} \boldsymbol{h}_j^l \parallel \textcolor{blue}{\boldsymbol{W}^{l,k}_{ya}} \boldsymbol{h}_i^l] )} {\sum_{k \in \mathcal{N}(j)}\exp(\textcolor{blue}{\boldsymbol{a}^T} [ \textcolor{blue}{\boldsymbol{W}^{l,k}_{xa}} \boldsymbol{h}_j^l \parallel \textcolor{blue}{\boldsymbol{W}^{l,k}_{ya}}  \boldsymbol{h}_k^l] )} \\
			&\boldsymbol{m}_{j, i, 1}^l = \parallel_{k=1}^K \delta(\alpha_{j, i}^k \textcolor{blue}{\boldsymbol{W}^{l,k}_v} \boldsymbol{h}_j^{l}) \\
			&\boldsymbol{m}_{j, i, 2}^l = \textcolor{blue}{\boldsymbol{W}^l_m} \boldsymbol{h}_j^{l} \\
			&\boldsymbol{m}_{j, i, 3}^l = \boldsymbol{h}_j^l
			\end{aligned}$
		\end{scriptsize} &
		$O(max(d_a, d_m) * K * d_{in})$ \\
		\bottomrule
	\end{tabular}
\end{footnotesize}
	\caption{Typical graph neural networks and their edge computation functions.$d_{in}$ and $d_{out}$ are dimensions of the input and output hidden feature vectors, respectively.
		Blue variables are model parameters to learn. For ChebNet, we report its GNN sub-layer.}
	\label{tab:gnn_overview_edge}
\end{table}

\begin{table}
	\hspace{-3em}
    \begin{footnotesize}
        \begin{tabular}{cp{20em}r}
		\toprule
		GNN &
		$\gamma$ &
		Complexity \\ \midrule
		ChebNet \cite{defferrad2016_chebnet} &
		$\boldsymbol{z}_i^{(k)} = 2\boldsymbol{s}^{(k)}_{i} - \boldsymbol{z}_i^{(k-2)}$&
		$O(d_{in})$ \\
		\textbf{GCN} &
		$\boldsymbol{h}_i^{l+1} = \textcolor{blue}{\boldsymbol{W}}^l  \boldsymbol{s}_i^{l}$ &
		$O(d_{in} * d_{out})$ \\
		AGCN &
		$\boldsymbol{h}_i^{l+1} = \textcolor{blue}{\boldsymbol{W}}^l  \boldsymbol{s}_i^{l}$ &
		$O(d_{in} * d_{out})$ \\
		GraphSAGE &	
		$\boldsymbol{h}_i^{l+1} =   \delta(\textcolor{blue}{\boldsymbol{W}}^l  [\boldsymbol{s}_i^{l} \parallel \boldsymbol{h}_i^l])$ &
		$O(d_{in} * d_{out})$ \\
		Neural FPs &
		$\boldsymbol{h}_i^{l+1} = \delta(\boldsymbol{h}_i^l + \textcolor{blue}{\boldsymbol{W}}^{\boldsymbol{N}_i}  \boldsymbol{s}_i^{l})$ &
		$O(d_{in} * d_{out})$ \\
		SSE &
		$\boldsymbol{h}_i^{l+1} = (1 - \alpha)  \boldsymbol{h}_i^l +\alpha    \delta(\textcolor{blue}{\boldsymbol{W}}^l_1 \delta(\textcolor{blue}{\boldsymbol{W}}^l_2), \boldsymbol{s}_i^l)$ &
		$O(d_{in} * d_{out})$ \\
		\textbf{GGNN} &
		\begin{scriptsize}
		$\begin{aligned}[t]
			& \textcolor{blue}{\boldsymbol{z}}_i^l = \delta ( \textcolor{blue}{\boldsymbol{W}}^z \boldsymbol{s}_i^l + \textcolor{blue}{\boldsymbol{b}}^{sz} + \textcolor{blue}{\boldsymbol{U}}^z \boldsymbol{h}_i^{l} + \textcolor{blue}{\boldsymbol{b}}^{hz}) \\ 
			&\boldsymbol{r}_i^l = \delta ( \textcolor{blue}{\boldsymbol{W}}^r \boldsymbol{s}_i^l+ \textcolor{blue}{\boldsymbol{b}}^{sr} +\textcolor{blue}{\boldsymbol{U}}^r \boldsymbol{h}_i^{l} + \textcolor{blue}{\boldsymbol{b}}^{hr}) \\ 
			&\boldsymbol{h}_i^{l+1} = tanh ( \textcolor{blue}{\boldsymbol{W}}^l \boldsymbol{s}_i^l + \textcolor{blue}{\boldsymbol{b}}^s + \textcolor{blue}{\boldsymbol{U}} ( \boldsymbol{r}_i^l \odot \boldsymbol{h}_i^{l} + \textcolor{blue}{\boldsymbol{b}}^h))) \\ 
			&\boldsymbol{h}_i^{l+1} = (1 - \textcolor{blue}{\boldsymbol{z}}_i^l) \odot \boldsymbol{h}_i^l +  \textcolor{blue}{\boldsymbol{z}}_i^l \odot \boldsymbol{h}_i^{l+1}
		\end{aligned}$
		\end{scriptsize}
		&
		$O(max(d_{in}, d_{out}) * d_{out})$ \\
		\textbf{GAT}&
		$\boldsymbol{h}_i^{l+1} = \boldsymbol{s}_i^l$ &
		$O(1)$ \\
		\textbf{GaAN} &
		\begin{scriptsize}
			$\begin{aligned}[t]
			&\boldsymbol{g}_i = \textcolor{blue}{\boldsymbol{W}}^l_g  [\boldsymbol{h}_i^{l} \parallel \boldsymbol{s}_{i, 2}^l \parallel \boldsymbol{s}_{i, 3}^l]  \\
			&\boldsymbol{h}_i^{l+1} = \textcolor{blue}{\boldsymbol{W}}^l_o [\boldsymbol{h}_i^l \parallel (\boldsymbol{g}_{i} \odot \boldsymbol{s}_{i, 3}^l) ]
			\end{aligned}$ 
		\end{scriptsize}&
		$O(max(d_{in} + K * d_v, 2 * d_{in} + d_m) d_{out})$ \\ \bottomrule
	\end{tabular}
\end{footnotesize}
	\caption{Typical graph neural networks and their vertex computation functions. $d_{in}$ and $d_{out}$ are dimensions of the input and output hidden feature vectors, respectively. Blue variables are model parameters to learn. For ChebNet, we report its GNN sub-layer.}
	\label{tab:gnn_overview_vertex}
\end{table}

@@

We classify the typical GNNs into four groups based on their edge/vertex complexity.

\begin{figure}
	\centering
	\includegraphics[width=0.7\columnwidth]{figs/illustration/GNN_complexity_quadrant.png}
	\caption{Complexity quadrant of typical GNNs. If two GNNs have the same big-O complexity, we further compare their complexity according to the number of parameters.}
\end{figure}

>>>>

\begin{table}
	\hspace{-10em}
    \begin{footnotesize}
        \begin{tabular}{ccp{8em}p{22em}r}
		\toprule
		GNN &
		Type &
		$\Sigma$ &
		$\phi$ &
		Complexity \\ \midrule
		ChebNet \cite{defferrad2016_chebnet} &
		Spectral &
		sum &
		$\boldsymbol{m}_{j, i}^{(k)} = e_{j, i}\boldsymbol{z}_j^{(k-1)}$ &
		$O(d_{in})$ \\
		\textbf{GCN} \cite{kipf2017_gcn} &
		Spectral &
		sum &
		$\boldsymbol{m}_{j, i}^l = e_{j, i} \boldsymbol{h}_j^l$ &
		$O(d_{in})$ \\
		AGCN &
		Spectral &
		sum &
		$\boldsymbol{m}_{j, i}^l = \tilde{e}_{j, i}^l \boldsymbol{h}_j^l$ &
		$O(d_{in})$ \\
		GraphSAGE &
		Non-spectral &
		mean/LSTM &
		$\boldsymbol{m}_{j, i}^l =  \boldsymbol{h}_j^l$ &
		$O(1)$ \\
		Neural FPs &
		Non-spectral &
		sum &
		$\boldsymbol{m}_{j, i}^l = \boldsymbol{h}_j^l$ &
		$O(d_{in})$ \\
		SSE &
		Recurrent &
		sum &
		$\boldsymbol{m}_{j, i}^l = [\boldsymbol{h}_i^{l} \parallel \boldsymbol{h}_j^l]$ &
		$O(1)$ \\
		\textbf{GGNN} &
		Gated &
		sum &
		$\boldsymbol{m}_{j, i} = \textcolor{blue}{\boldsymbol{W}^l} \boldsymbol{h}_j^l$ &
		$O(d_{in} * d_{out})$ \\
		\textbf{GAT}&
		Attention &
		sum  &
		\begin{scriptsize}
			$\begin{aligned}[t]
			& \alpha_{j, i}^k = \frac {\exp(LeakyReLU(\textcolor{blue}{\boldsymbol{a}^T} [ \textcolor{blue}{\boldsymbol{W}^{l,k}}  \boldsymbol{h}_j^l \parallel \textcolor{blue}{\boldsymbol{W}^{l,k}}  \boldsymbol{h}_j^l] ))} {\sum_{k \in \mathcal{N}(i)}\exp(LeakyReLU(\textcolor{blue}{\boldsymbol{a}^T} [ \textcolor{blue}{\boldsymbol{W}^{l,k}}  \boldsymbol{h}_j^l \parallel \textcolor{blue}{\boldsymbol{W}^{l,k}}  \boldsymbol{h}_k^l] ))} \\
			&  \boldsymbol{m}_{j, i}^l = \parallel_{k=1}^K \delta(\alpha_{j, i}^k \textcolor{blue}{\boldsymbol{W}^{l,k}} \boldsymbol{h}_j^{l})
			\end{aligned}$
		\end{scriptsize} &
		$O(K * d_{in} * d_{out})$ \\
		\textbf{GaAN} &
		Attention &
		sum,max,mean &
		\begin{scriptsize}
			$\begin{aligned}[t]
			&\alpha_{j, i}^k = \frac {\exp(\textcolor{blue}{\boldsymbol{a}^T} [ \textcolor{blue}{\boldsymbol{W}^{l, k}_{xa}} \boldsymbol{h}_j^l \parallel \textcolor{blue}{\boldsymbol{W}^{l,k}_{ya}} \boldsymbol{h}_i^l] )} {\sum_{k \in \mathcal{N}(j)}\exp(\textcolor{blue}{\boldsymbol{a}^T} [ \textcolor{blue}{\boldsymbol{W}^{l,k}_{xa}} \boldsymbol{h}_j^l \parallel \textcolor{blue}{\boldsymbol{W}^{l,k}_{ya}}  \boldsymbol{h}_k^l] )} \\
			&\boldsymbol{m}_{j, i, 1}^l = \parallel_{k=1}^K \delta(\alpha_{j, i}^k \textcolor{blue}{\boldsymbol{W}^{l,k}_v} \boldsymbol{h}_j^{l}) \\
			&\boldsymbol{m}_{j, i, 2}^l = \textcolor{blue}{\boldsymbol{W}^l_m} \boldsymbol{h}_j^{l} \\
			&\boldsymbol{m}_{j, i, 3}^l = \boldsymbol{h}_j^l
			\end{aligned}$
		\end{scriptsize} &
		$O(max(d_a, d_m) * K * d_{in})$ \\
		\bottomrule
	\end{tabular}
\end{footnotesize}
	\caption{Typical graph neural networks and their edge computation functions.$d_{in}$ and $d_{out}$ are dimensions of the input and output hidden feature vectors, respectively.
		Blue variables are model parameters to learn. For ChebNet, we report its GNN sub-layer.}
	\label{tab:gnn_overview_edge}
\end{table}

\begin{table}
	\hspace{-3em}
    \begin{footnotesize}
        \begin{tabular}{cp{20em}r}
		\toprule
		GNN &
		$\gamma$ &
		Complexity \\ \midrule
		ChebNet \cite{defferrad2016_chebnet} &
		$\boldsymbol{z}_i^{(k)} = 2\boldsymbol{s}^{(k)}_{i} - \boldsymbol{z}_i^{(k-2)}$&
		$O(d_{in})$ \\
		\textbf{GCN} \cite{kipf2017_gcn} &
		$\boldsymbol{h}_i^{l+1} = \textcolor{blue}{\boldsymbol{W}}^l  \boldsymbol{s}_i^{l}$ &
		$O(d_{in} * d_{out})$ \\
		AGCN &
		$\boldsymbol{h}_i^{l+1} = \textcolor{blue}{\boldsymbol{W}}^l  \boldsymbol{s}_i^{l}$ &
		$O(d_{in} * d_{out})$ \\
		GraphSAGE &	
		$\boldsymbol{h}_i^{l+1} =   \delta(\textcolor{blue}{\boldsymbol{W}}^l  [\boldsymbol{s}_i^{l} \parallel \boldsymbol{h}_i^l])$ &
		$O(d_{in} * d_{out})$ \\
		Neural FPs &
		$\boldsymbol{h}_i^{l+1} = \delta(\boldsymbol{h}_i^l + \textcolor{blue}{\boldsymbol{W}}^{\boldsymbol{N}_i}  \boldsymbol{s}_i^{l})$ &
		$O(d_{in} * d_{out})$ \\
		SSE &
		$\boldsymbol{h}_i^{l+1} = (1 - \alpha)  \boldsymbol{h}_i^l +\alpha    \delta(\textcolor{blue}{\boldsymbol{W}}^l_1 \delta(\textcolor{blue}{\boldsymbol{W}}^l_2), \boldsymbol{s}_i^l)$ &
		$O(d_{in} * d_{out})$ \\
		\textbf{GGNN} &
		\begin{scriptsize}
		$\begin{aligned}[t]
			& \textcolor{blue}{\boldsymbol{z}}_i^l = \delta ( \textcolor{blue}{\boldsymbol{W}}^z \boldsymbol{s}_i^l + \textcolor{blue}{\boldsymbol{b}}^{sz} + \textcolor{blue}{\boldsymbol{U}}^z \boldsymbol{h}_i^{l} + \textcolor{blue}{\boldsymbol{b}}^{hz}) \\ 
			&\boldsymbol{r}_i^l = \delta ( \textcolor{blue}{\boldsymbol{W}}^r \boldsymbol{s}_i^l+ \textcolor{blue}{\boldsymbol{b}}^{sr} +\textcolor{blue}{\boldsymbol{U}}^r \boldsymbol{h}_i^{l} + \textcolor{blue}{\boldsymbol{b}}^{hr}) \\ 
			&\boldsymbol{h}_i^{l+1} = tanh ( \textcolor{blue}{\boldsymbol{W}}^l \boldsymbol{s}_i^l + \textcolor{blue}{\boldsymbol{b}}^s + \textcolor{blue}{\boldsymbol{U}} ( \boldsymbol{r}_i^l \odot \boldsymbol{h}_i^{l} + \textcolor{blue}{\boldsymbol{b}}^h))) \\ 
			&\boldsymbol{h}_i^{l+1} = (1 - \textcolor{blue}{\boldsymbol{z}}_i^l) \odot \boldsymbol{h}_i^l +  \textcolor{blue}{\boldsymbol{z}}_i^l \odot \boldsymbol{h}_i^{l+1}
		\end{aligned}$
		\end{scriptsize}
		&
		$O(max(d_{in}, d_{out}) * d_{out})$ \\
		\textbf{GAT}&
		$\boldsymbol{h}_i^{l+1} = \boldsymbol{s}_i^l$ &
		$O(1)$ \\
		\textbf{GaAN} &
		\begin{scriptsize}
			$\begin{aligned}[t]
			&\boldsymbol{g}_i = \textcolor{blue}{\boldsymbol{W}}^l_g  [\boldsymbol{h}_i^{l} \parallel \boldsymbol{s}_{i, 2}^l \parallel \boldsymbol{s}_{i, 3}^l]  \\
			&\boldsymbol{h}_i^{l+1} = \textcolor{blue}{\boldsymbol{W}}^l_o [\boldsymbol{h}_i^l \parallel (\boldsymbol{g}_{i} \odot \boldsymbol{s}_{i, 3}^l) ]
			\end{aligned}$ 
		\end{scriptsize}&
		$O(max(d_{in} + K * d_v, 2 * d_{in} + d_m) d_{out})$ \\ \bottomrule
	\end{tabular}
\end{footnotesize}
	\caption{Typical graph neural networks and their vertex computation functions. $d_{in}$ and $d_{out}$ are dimensions of the input and output hidden feature vectors, respectively. Blue variables are model parameters to learn. For ChebNet, we report its GNN sub-layer.}
	\label{tab:gnn_overview_vertex}
\end{table}

@@

We classify the typical GNNs into four quadrants based on their edge/vertex complexity as shown in \figurename~\ref{fig:gnn_complexity_quadrant}. We pick GCN, GGNN, GAT and GaAN as the representative GNNs of the four quadrants.

\begin{figure}
	\centering
	\includegraphics[width=0.7\columnwidth]{figs/illustration/GNN_complexity_quadrant.png}
	\caption{Complexity quadrants of typical GNNs. If two GNNs have the same complexity under the Big-O notation, we further compare their complexity according to how many parameters they need to train.}
	\label{fig:gnn_complexity_quadrant}
\end{figure}

\textbf{GCN} \cite{kipf2017_gcn} (low vertex \& edge computational complexity): Graph convolution network (GCN) is the first-order approximation of the spectral-based graph convolutions.
%By combining the advantages of both the spectral-based and the spatial-based convolution, it can learn the representation of the local topological structure and vertex features efficiently.
It has only one parameter to learn at each layer, i.e. the weight matrix $\boldsymbol{W}^l$ in $\gamma$.
A GCN graph neuron can be expressed as $\boldsymbol{h}^{l+1}_i = \boldsymbol{W}^l\sum_{v_j \in \mathcal{N}(v_i)}{e_{j,i}\boldsymbol{h}^l_j}$, where $e_{j,i}$ is the normalized weight of the edge $(v_j, v_i)$.
%In the matrix view, $\boldsymbol{H}^{l+1} = (\boldsymbol{A}\boldsymbol{H}^l)\boldsymbol{W}^l$.
According to the associative law of the matrix multiplication, $\boldsymbol{h}^{l+1}_i = \sum_{v_j \in \mathcal{N}(v_i)}{e_{j,i}\boldsymbol{W}^l\boldsymbol{h}^l_j}$.
Since the dimension of $\boldsymbol{h}^{l+1}_i$ is usually smaller than $\boldsymbol{h}^l_i$ in practical GCNs, the implementation of GCN \footnote{The implementation of GCN in PyTorch Geometric: \url{ https://github.com/rusty1s/pytorch_geometric/blob/1.5.0/torch_geometric/nn/conv/gcn_conv.py}} chooses to first conduct the vertex computation $\hat{\boldsymbol{h}}^l_j = \boldsymbol{W}^l\boldsymbol{h}^l_j$ for each vertex $v_j$ and then conduct the edge computation $\boldsymbol{h}^{l+1}_i=\sum_{v_j\in\mathcal{N}(v_i)}{\hat{\boldsymbol{h}}^l_j}$.
As $\hat{\boldsymbol{h}}^l_j$ has the same dimension as $\boldsymbol{h}^{l+1}_i$, the implementation significantly reduces the  computation cost of the edge computation.
